---
title: "Assignment #9 – Visualization in R (Base, lattice, ggplot2)"
author: "Gabriel Myles"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.path = "figs/")
```

## Dataset

We use **mtcars** from the **datasets** package (Rdatasets collection).

```{r}
library(lattice)
library(ggplot2)
data("mtcars", package = "datasets")
mtcars$cyl_f <- factor(mtcars$cyl)
mtcars$gear_f <- factor(mtcars$gear)
head(mtcars)
```

## 1) Base R Graphics

### A. Scatterplot with OLS line

```{r}
plot(mtcars$wt, mtcars$mpg,
     pch = 19, col = "steelblue",
     xlab = "Weight (1000 lbs)", ylab = "Miles per Gallon (MPG)",
     main = "Base R: MPG vs Weight")
abline(lm(mpg ~ wt, data = mtcars), col = "firebrick", lwd = 2)
legend("topright", legend = c("Cars", "OLS fit"),
       pch = c(19, NA), lty = c(NA, 1), col = c("steelblue", "firebrick"), bty = "n")
```

**Interpretation.** Heavier cars tend to have lower MPG; the linear trend is clearly negative.

### B. Histogram of MPG

```{r}
hist(mtcars$mpg,
     breaks = 10, col = "gray85", border = "white",
     main = "Base R: Distribution of MPG", xlab = "MPG")
rug(mtcars$mpg)
```

**Interpretation.** MPG is roughly unimodal with a slight right tail.

## 2) lattice Graphics

### A. Small-multiple scatter by cylinders

```{r}
xyplot(mpg ~ wt | cyl_f, data = mtcars,
       pch = 16,
       xlab = "Weight (1000 lbs)", ylab = "MPG",
       main = "lattice: MPG vs Weight by Cylinders",
       panel = function(x, y, ...) {
         panel.grid(h = -1, v = -1)
         panel.xyplot(x, y, ...)
         panel.lmline(x, y, col = "firebrick", lwd = 2)
       })
```

### B. Boxplot of MPG by gears

```{r}
bwplot(mpg ~ gear_f, data = mtcars,
       ylab = "MPG", xlab = "Gears",
       main = "lattice: MPG by Number of Gears")
```

## 3) ggplot2 Graphics

### A. Scatter with smoothing (color = cylinders)

```{r}
ggplot(mtcars, aes(wt, mpg, color = cyl_f)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "ggplot2: MPG vs Weight (with linear trend)",
       x = "Weight (1000 lbs)", y = "MPG", color = "Cylinders") +
  theme_minimal(base_size = 13)
```

### B. Faceted MPG histogram by cylinders

```{r}
ggplot(mtcars, aes(mpg)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~ cyl_f, nrow = 1) +
  labs(title = "ggplot2: MPG Distribution by Cylinders", x = "MPG", y = "Count") +
  theme_classic(base_size = 13)
```

## Discussion

**Workflow differences.** Base graphics build plots by calling functions (`plot()`, `hist()`) and adding layers (like `abline()`). **lattice** encourages *formulas* and *conditioning* (e.g., `y ~ x | group`) and returns plot objects that are printed. **ggplot2** uses a grammar of layered components (`geom_`, `aes()`, `+`) and provides consistent theming and scales.

**Publication quality vs control.** ggplot2 gave the cleanest results with minimal code and easiest theming. lattice excels at small multiples. Base is flexible but requires more manual styling.

**Challenges/surprises.** Remembering where to add elements differs: `abline()` in base, `panel.lmline()` inside a lattice `panel`, and `geom_smooth(method="lm")` in ggplot2.

---

*End of report.*
